---
title: "Final Project"
author: "Jhony Sanchez & Rodrigo Malagón"
date: "2024-07-11"
output: github_document
---

```{r}
library(keras)
library(tensorflow)
library(tfdatasets)
library(purrr)
library(ggplot2)
library(rsample)
library(stars)
library(raster)
library(reticulate)
library(mapview)
```

# I. Pre-processing

We begin our workflow with a preprocessing stage whose main objective is to obtain a labelled dataset consisting of tile multispectral images of the area of study and their corresponding labelling masks. The workflow was conducted on QGis.

After the images were collected from the open data [Xpress platform](https://xpress.maxar.com/?lat=0&lon=0&zoom=3.0) by Maxar, we assemble a mosaic of the study area. The mosaic was clipped into a single regular square raster. 

Afterwards, we break the area of study into 12 raster images of uniform size to break the optimize furthcoming preprocessing stages with these smaller inputs.

The `GenericRegionMerging` segmentation tool from the OTB toolbox is then used per image to create uniform patches accross each image and thus facilitate the labelling procedure. To this end, the `scale factor` was set to $350$ and the spectral and spatial homogeneity parameters to the default values of $0.5$. 

After this, the visual identification and manual labelling of _flooded_ segments was performed. Then, we transform the labelled vectors into binary images and perform a tiling of the whole dataset into more than $8500$ images of size $128\times 128$.

# II. Modelling

In this main stage of the project, we follow the general workflow exposed in the tutorial [Introduction to Deep Learning in R for the Analysis of UAV-based Remote Sensing Data](https://dachro.github.io/ogh_summer_school_2020/Tutorial_DL_UAV.html#from_image_to_pixel-by-pixel_classification). As key changes to this general workflow, we adapt some processing steps to the multispectral data structure of our project and perform a stratification of masks to enhance the dataset split into training and testing sets.

## II.1 Dataset preparation

In this stage, we begin defining some helping functions to process the tile imagery in preparation to the tensor dataset creation. We stress that though this process may be carried in general after the creation of the tensor dataset, we perform this processing due to the lack of straightforward reading capabilities of `.tiff` files by the `keras` API.

```{r}
# Processing functions
extract_matrix <- function(file){
  raster <- read_stars(file)
  matrix <- raster[[1]]
  matrix
}

extract_mask <- function(file){
  raster <- read_stars(file)
  matrix <- raster[,,][[1]]
  matrix
}


normalize_matrix <- function(matrix){
  matrix/max(matrix)
}
```

Due to processing limitations, we select a subset of $5000$ labelled tiles to carry out the CNN learning stage:

```{r}
path_to_imgs <- './Data/10300100FAA4CA00/tiles_image/'
path_to_masks <- './Data/10300100FAA4CA00/tiles_labels/'

(files <- data.frame(
  img_path = list.files(path_to_imgs, full.names = TRUE, pattern = "*.TIF$"),
  mask_path = list.files(path_to_masks, full.names = TRUE, pattern = "*.TIF$")
))
files<-files[1:5000,]
```


We then extract the multispectral images and normalize their values to work with features in the $[0,1]$ interval.

```{r}
files$img <- lapply(files$img_path,extract_matrix)

files$mask <- lapply(files$mask_path,extract_mask)


files$img <- lapply(files$img,normalize_matrix)
head(files)
```

We filter the tiles with a $128\times 128-$size to produce a uniform dataset to be fed to the CNN. Furthermore, we enrich our dataset with an extracted *mask weight* that ranks masks from 0 to 10 in terms of quantity of 1's.

```{r}
# Data size filtering 
files$dim_x <- lapply(files$img,function(x){dim(x)['x']})|> unlist()
files <- files[files[,'dim_x'] == 128,] 
files$dim_y <- lapply(files$img,function(x){dim(x)['y']}) |> unlist()
files <- files[files[,'dim_y'] == 128,]

# Mask weight creation
files$mask_weight <- unlist(lapply(files$mask,function(x){as.integer(10*round(mean(x),1))}))
      #integer type enforced to be passed as strata parameter in the next step 
```

We perform a data split with a stratification of our dataset using the *mask weight* created in the last step.

```{r}
# Dataset split
files_split <- initial_split(files[c('img','mask','mask_weight')],prop = 0.8,strata=mask_weight)

training_dataset <- tensor_slices_dataset(training(files_split)[c('img','mask')])
      #we leave the mask weight out of the tensor dataset, as this component cannot be fed to the CNN
testing_dataset <- tensor_slices_dataset(testing(files_split)[c('img','mask')])
```

One further change of type to single-precision floating-point is performed to our dataset to ensure a lower memory demand.

```{r}
training_dataset <- dataset_map(training_dataset, function(.x)
   list_modify(.x, img = tf$image$convert_image_dtype(.x$img, dtype = tf$float32),
               mask = tf$image$convert_image_dtype(.x$mask, dtype = tf$float32)))

testing_dataset <- dataset_map(testing_dataset, function(.x)
   list_modify(.x, img = tf$image$convert_image_dtype(.x$img, dtype = tf$float32),
                mask = tf$image$convert_image_dtype(.x$mask, dtype = tf$float32)))
```

Finally, we shuffle the training dataset and batch both the training and testing datasets.

```{r}
dataset_iterator <- as_iterator(training_dataset)
dataset_list <- iterate(dataset_iterator)
(training_dataset_size <- length(dataset_list))

training_dataset <- dataset_shuffle(training_dataset, buffer_size = training_dataset_size)
training_dataset <- dataset_batch(training_dataset, 10L)
training_dataset <- dataset_map(training_dataset, unname)

testing_dataset <- dataset_batch(testing_dataset, 10L)
testing_dataset <- dataset_map(testing_dataset, unname)
```

# II.2 CNN model

In this stage we define the CNN architecture that we use for the modelling. To this end, we use the U-net architecture exposed in the tutorial [Introduction to Deep Learning in R for the Analysis of UAV-based Remote Sensing Data](https://dachro.github.io/ogh_summer_school_2020/Tutorial_DL_UAV.html#from_image_to_pixel-by-pixel_classification). The meaningful change performed to this architecture is the adaptation of the `input shape` of the CNN to our tiles of size $128\times 128$.

```{r}
## we start with the "contratcing path"##
# input
input_tensor <- layer_input(shape = c(128,128,8))

#conv block 1
unet_tensor <- layer_conv_2d(input_tensor,filters = 64,kernel_size = c(3,3), padding = "same",activation = "relu")
conc_tensor2 <- layer_conv_2d(unet_tensor,filters = 64,kernel_size = c(3,3), padding = "same",activation = "relu")
unet_tensor <- layer_max_pooling_2d(conc_tensor2)

#conv block 2
unet_tensor <- layer_conv_2d(unet_tensor,filters = 128,kernel_size = c(3,3), padding = "same",activation = "relu")
conc_tensor1 <- layer_conv_2d(unet_tensor,filters = 128,kernel_size = c(3,3), padding = "same",activation = "relu")
unet_tensor <- layer_max_pooling_2d(conc_tensor1)

#"bottom curve" of unet
unet_tensor <- layer_conv_2d(unet_tensor,filters = 256,kernel_size = c(3,3), padding = "same",activation = "relu")
unet_tensor <- layer_conv_2d(unet_tensor,filters = 256,kernel_size = c(3,3), padding = "same",activation = "relu")

##  this is where the expanding path begins ##

# upsampling block 1
unet_tensor <- layer_conv_2d_transpose(unet_tensor,filters = 128,kernel_size = c(2,2),strides = 2,padding = "same") 
unet_tensor <- layer_concatenate(list(conc_tensor1,unet_tensor))
unet_tensor <- layer_conv_2d(unet_tensor, filters = 128, kernel_size = c(3,3),padding = "same", activation = "relu")
unet_tensor <- layer_conv_2d(unet_tensor, filters = 128, kernel_size = c(3,3),padding = "same", activation = "relu")

# upsampling block 2
unet_tensor <- layer_conv_2d_transpose(unet_tensor,filters = 64,kernel_size = c(2,2),strides = 2,padding = "same")
unet_tensor <- layer_concatenate(list(conc_tensor2,unet_tensor))
unet_tensor <- layer_conv_2d(unet_tensor, filters = 64, kernel_size = c(3,3),padding = "same", activation = "relu")
unet_tensor <- layer_conv_2d(unet_tensor, filters = 64, kernel_size = c(3,3),padding = "same", activation = "relu")

# output
unet_tensor <- layer_conv_2d(unet_tensor,filters = 1,kernel_size = 1, activation = "sigmoid")

# combine final unet_tensor (carrying all the transformations applied through the layers) 
# with input_tensor to create model

unet_model <- keras_model(inputs = input_tensor, outputs = unet_tensor)
```

We then compile and run the model with the `RMSprop` weight optimization algorithm and an optimal learning rate set to $1\times 10^{-4}$. The loss function minimized by the model learning is set to `Binary crossentropy` function and the metrics for the evaluation of the model as the `metric binary accuracy`, as our model predicts on two classes across each tile.

```{r}
# Model compilation
compile(
  unet_model,
  optimizer = optimizer_rmsprop(learning_rate = 1e-4),
  loss = "binary_crossentropy",
  metrics = c(metric_binary_accuracy)
  )

```

```{r}
# Model learning and evaluation metrics
diagnostics <- fit(unet_model,
               training_dataset,
               epochs = 10,
               validation_data = testing_dataset)

plot(diagnostics)
```

The model can be locally saved for further use:

```{r}
#save_model_hdf5(unet_model,filepath = "./FloodsModel.h5")
```

## II.3 Model prediction

In this stage we leverage the learned model to visually evaluate the class prediction on the testing tiles.

```{r}
flood_model <- load_model_hdf5("./FloodsModel.h5")
```

After loading the previously saved model, we select an image to be tested in this prediction stage. To assemble a visual comparison between the expected and the actual prediction (mask vs. predicted mask), we use of the `magick` library to stack these images.

```{r}
pred <- predict(object = flood_model,testing_dataset)
image_num <- 925

pred_img <- as.raster(pred[image_num,,,]) |> magick::image_read()
mask <- as.raster(testing(files_split)[image_num,c('mask')][[1]]) |> magick::image_read()
img <- as.raster(testing(files_split)[image_num,c('img')][[1]][,,4:2]) |> magick::image_read()


out <- magick::image_append(c(
  magick::image_append(mask, stack = TRUE),
  magick::image_append(img, stack = TRUE), 
  magick::image_append(pred_img, stack = TRUE)
)
)

plot(out)
```
